---
title: 'Lab 9: Malaria'
author: "Present Group Member's Names Go Here!"
subtitle: "Two-way Chi-square Test"
output: pdf_document
editor: visual
---

```{r setup}
#| include: false
## Load in the packages necessary for our analysis: tidyverse
knitr::opts_chunk$set(message = FALSE, 
                      fig.align = "center", 
                      echo = TRUE,
                      error = TRUE)

## Load in the packages necessary for our analysis: tidyverse
library(tidyverse)
library(infer)
library(janitor)
```

An article that appeared in the journal Lancet in May of 2021 (Datoo, et al.) described a study that investigated a potential vaccine that might protect children against malaria. Researchers recruited children between the ages of 5 and 17 months in Burkina Faso, a country in western Africa, as participants. The children were randomly assigned into one of three groups: one group received a large dose of the potential vaccine, another group received a small dose, and a third group received a placebo. Researchers observed the children for the next 18 months, keeping track of whether or not the child developed malaria. Researchers hoped, of course, that children who received a vaccine would be less likely to develop malaria than children who received a placebo.

## Setup

```{r, load-data}
malaria_data <- read_csv("data/malaria_data.csv")
malaria_data
```

**1. What is the observational unit for the study?**

answer goes here...

**2. Which is the explanatory variable and which is the response variable? Specify the variable types and levels/units**

-   Explanatory:

-   Response:

**3. Should you do a chi-square test of independence or test of homogeneity? Justify how you know.**

answer goes here...

## Visualize & Summarize the Data

4.  Fill in the code below to make a filled bar plot of the data. What do you see in your filled-bar plot?

```{r filled-barplot}
ggplot(data = __________,
       mapping = aes(x = _________, fill = _______)) +
  geom_bar(position = "______") +
  labs(x = "____________",
       y = "____________")
```

5.  Why do we want to make a filled bar plot over a stacked or dodged bar plot?

    answer goes here...

6.  Finish the code below to make a two-way contingency table. Note: we typically want our explanatory variable to be indicated in the columns.

```{r two-way-table}
malaria_data %>% 
   count(_____________, _____________)%>% 
   pivot_wider(names_from = _______________, 
               values_from = n) %>% 
   adorn_totals(where = c("row", "col"))
```

**7. What proportion of children who received the high dose vaccine contracted malaria?**

answer goes here...

**8. What proportion of children who received the low dose vaccine contracted malaria?**

answer goes here...

**9. What proportion of children who received the placebo contracted malaria?**

answer goes here...

## Theory-based Chi-square

**10. Write the null and alternative hypothesis for this study in words.**

-   Null:

-   Alternative:

Recall, in order for the $\chi^2$ distribution to be a good approximation of the true sampling distribution, we need to verify two conditions:

-   The observations are independent
-   We have a "large enough" sample size
    -   This is checked by verifying there are at least 5 expected counts in each cell

11. Is the independent observation condition met? Justify your answer.

answer goes here...

The equation for calculating expected counts is:

$$\frac{\text{row } i \text{ total count}\times \text{column } j \text{ total count}}{\text{total count}}$$ Our table is only a 2 x 3 table, but what if you had a 6 x 8 or worse, 20 x 42 table? Checking each cell's expected count would be very tedious. We only need to check the cell which will have the smallest expected count.

12. Which row in your two-way contingency table from #6 has the smallest total count?

    answer goes here...

13. Which column in your two-way contingency table from #6 has the smallest total count?

    answer goes here...

**14. Using the equation above calculate the expected count for the cell in the row and column you specified in #10 and #11.**

answer goes here...

**15. Is the "large enough" sample size condition met?**

answer goes here...

**16. Can we use the** $\chi^2$ distribution to approximate the true sampling distribution? Which conditions were necessary to have been checked?

17. Fill in the code below to perform a theory based Chi-square test of independence.

```{r chi-square-test-theory}
chisq_test(x = malaria_data, 
           response = ________________, 
           explanatory = ________________)
```

18. What conclusion would you reach based on your results? *Make sure to address (1) Chi-square test statistic and associated degrees of freedom, (2) p-value, (3)* $\alpha$ threshold, (4) your decision about the null hypothesis, and (5) your conclusion in context of the data.

answer goes here...give me a nice paragraph!

## Simulated Chi-square

What if our conditions had not been met? We would have needed to use a simulation based approach. Let's walk through what this would look like.

**19. First, we need to calculate the observed chi-square test statistic from our data. We did this by hand during the activities, but it can be tedious so let's make R do it for us. Fill in the code below to calculate your observed chi-square test statistic.**

```{r observed-chisq}
obs_xsq <- malaria_data %>%
  specify(response = _____________________,
          explanatory = __________________) %>% 
  calculate(stat = "_______")

obs_xsq
```

20. What value of the observed chi-square test statistic did you calculate above? Where have we previously seen this?

answer goes here...

**21. Now we need to generate what the sampling distribution would look like if the null were true (aka null distribution of our chi-square statistics). Fill in the code below to generate and visualize the null distribution.**

```{r null-distribution}
null_dist <- malaria_data %>%
  specify(response = Malaria, 
          explanatory = Vaccine) %>% 
  hypothesize(null = "______________") %>%
  generate(reps = 1000, type = "__________") %>%
  calculate(stat = "Chisq")

visualize(data = null_dist, 
          method = "simulation")
```

**22. Once we have our null distribution, we can use this to calculate our simulated p-value.**

```{r simulated-pvalue}
get_pvalue(x = null_dist,
           obs_stat = obs_xsq,
           direction = "__________")
```

23. What conclusion would you reach with the simulated chi-square test? Does this differ from your answer in #18?

answer goes here...
